---
title: "Outlier Detection and Robust Linear Regression"
pdf-engine: xelatex
documentclass: article
date: today
date-format: "MMMM YYYY"
header-includes:
    - \usepackage{amsmath}
    - \usepackage{booktabs}
    - \usepackage{algorithm}
    - \usepackage{algpseudocode}
    - \usepackage{float}
fig-pos: "H"
geometry: margin=1in
execute:
    echo: false
    warning: false
    message: false
bibliography: references.bib
csl: cardiff-university-harvard.csl
link-citations: true
toc: true
toc-depth: 2
number-sections: true
fontsize: 11pt

format:
    pdf:
        colorlinks: true
        toccolor: black
        linkcolor: blue
        citecolor: blue
        urlcolor: blue
abstract: |
    Test
---

# Introduction

## Test

# Simple Outlier Detection and Preliminaries

# M-Estimators

Many models applying statistics or classification are highly sensitive to outliers. Even one extreme outlier can skew estimates, resulting in unreliable results. Therefore, we must ensure stable estimation methods to manoeuvre around these problematic outliers.

One of the main estimators in robust statistics is the M-estimator, meaning maximum likelihood–type estimator. These provide a flexible method for estimating parameters.

They are defined as solutions to optimisation problems by choosing a loss function. Let $x_1, \dots, x_n$ be observations and let $\rho(\cdot)$ be a loss function. An M-estimator of parameter $\theta$ is defined as

$$
\hat{\theta} = \arg\min_\theta \sum_{i=1}^n \rho(x_i - \theta).
$$

We obtain the estimating equations by differentiating the objective function. Let

$$
\psi(u) = \rho'(u).
$$

Then the estimator satisfies

$$
\sum_{i=1}^n \psi(x_i - \hat{\theta}) = 0.
$$

Different choices of $\rho$ lead to different estimators. For example, absolute loss yields the sample median, while squared loss returns the sample mean.

M-estimators can be used for an array of parameters including location, scale, and regression coefficients. Location M-estimators provide alternatives to the sample mean, scale estimators provide robustness in measuring dispersion, and in this report we focus on regression M-estimators, which can be interpreted as weighted least squares.

## Regression M-Estimators

For robust regression we consider the linear regression model

$$
y_i = x_i^T \beta + \varepsilon_i, \quad i = 1, \dots, n
$$

where $x_i$ is a vector of predictors and $\beta$ is the vector of regression coefficients. The residual corresponding to a parameter value $\beta$ is

$$
r_i(\beta) = y_i - x_i^T \beta.
$$

Regression M-estimators are defined by minimizing a robust objective function of the standardized residuals

$$
\hat{\beta} =
\arg\min_\beta
\sum_{i=1}^n
\rho \left( \frac{r_i(\beta)}{s} \right),
$$ {#eq-m-estimator-minimiser}

where $s$ is a scale estimate used to standardize the residuals.

The corresponding estimating equations are

$$
\sum_{i=1}^n
\psi \left( \frac{r_i(\beta)}{s} \right) x_i = 0,
\quad \text{where} \quad \psi = \rho'.
$$

These equations can also be written in the form

$$
\sum_{i=1}^n w_i r_i(\beta)x_i = 0,
$$

with weights

$$
w_i =
\frac{\psi\left(\frac{r_i(\beta)}{s}\right)}
{\frac{r_i(\beta)}{s}}.
$$

Since many quantities depend on one another, the estimation procedure is iterative. The residuals depend on the estimate of $\beta$, the weights $w_i$ depend on the residuals through the $\psi$ function, and the updated estimate of $\beta$ depends on the weights.

Typically, the estimation is performed using iteratively reweighted least squares (IRLS). We start from an initial estimate of $\beta$, compute the residuals, determine the weights from the $\psi$ function, and then compute a weighted least squares fit. This procedure is repeated until convergence.

A scale estimate $s$ is used to standardize the objective function. The scale can either be fixed or estimated jointly with the regression parameters. It is important that $s$ is robust, since a non-robust scale could become very large when outliers are present. In that case the standardized residuals $\frac{r_i(\beta)}{s}$ become artificially small and the weighting mechanism no longer reduces the influence of extreme observations.

A common choice for the initial scale estimate is the normalized median absolute deviation (MADN).

The behaviour of the estimator is determined by the choice of loss function $\rho$, its derivative $\psi$, and the associated weight function $w$.

```{r}
#| label: library-setup
library(knitr)
library(kableExtra)
library(tidyverse)
library(glue)
library(latex2exp)
library(patchwork)

source("../functions.R")
```
```{r}
#| label: fig-psi-rho-weight-functions
#| fig-cap: Plots of $\rho$, $\psi$ and weight functions of Huber and Bisquare M-estimators.
#| fig-width: 10
#| fig-height: 6

k <- 1.548
x_vals <- seq(-3, 3, length.out = 100)
plotting_df <- data.frame(
    x = x_vals, h_rho = huber_rho(x_vals, k),
    h_psi = huber_psi(x_vals, k = k),
    bi_rho = bisquare_rho(x_vals, k = k),
    bi_psi = bisquare_psi(x_vals, k = k),
    h_weight = huber_weight(x_vals, k = k),
    bi_weight = bisquare_weight(x_vals, k = k)
)
# maybe the following can be condensed...

p1 <- ggplot(plotting_df) +
    geom_line(aes(x = x, y = h_rho), color = "blue") +
    geom_line(aes(x = x, y = bi_rho), color = "red") +
    labs(x = TeX("$x$"), y = TeX("$\\rho(x)$")) +
    theme_minimal()

p2 <- ggplot(plotting_df) +
    geom_line(aes(x = x, y = h_psi, color = "Huber")) +
    geom_line(aes(x = x, y = bi_psi, color = "Bisquare")) +
    scale_color_manual(values = c("Huber" = "blue", "Bisquare" = "red")) +
    labs(x = TeX("$x$"), y = TeX("$\\psi(x)$"), color = "Estimator") +
    theme_minimal()

p3 <- ggplot(plotting_df) +
    geom_line(aes(x = x, y = h_weight), color = "blue") +
    geom_line(aes(x = x, y = bi_weight), color = "red") +
    labs(x = TeX("$x$"), y = "Weight") +
    theme_minimal()

p1 / p2 / p3 # this is from patchwork
```


Two common choices are the Huber and Bisquare functions, see @fig-psi-rho-weight-functions for their plots. The Huber function limits the influence of large residuals but never assigns a weight of zero, meaning extreme observations still have a small influence on the fit. In contrast, the bisquare function assigns zero weight to sufficiently large residuals, effectively removing these observations from the fitting procedure. M-estimators which satisfy this are called *redescending* and they are the most popular for linear regression, despite them not necessarily producing a unique solution of @eq-m-estimator-minimiser. 

While regression M-estimators improve robustness compared to ordinary least squares, the initial fit can still be sensitive to outliers. Even with a robust initial scale, large outliers may influence the early iterations of the algorithm. Consequently, regression M-estimators may not be sufficiently robust in all situations. This motivates the development of more robust procedures such as S-estimators, which are discussed in the following section.

# Regression MM-Estimators and Fast-S

MM-estimators were introduced by @10.1214/aos/1176350366 and work on a two-step process of starting with an S-estimator to robustly estimate scale, followed by M-estimation with fixed scale coming from the previous step. The S-estimator also provides a starting point for estimating $\beta$ in the second step. They are able to inherit the high breakdown point of the S-estimator, while also keeping the high normal efficiency of an M-estimator. The M-estimation step follows @eq-m-estimator-minimiser as in the previous section.

Let $p$ denote the number of predictors in a regression problem. Computing an S-estimator is computationally challenging, but there are viable approximations for $p$ small. We have implemented the *Fast-S* algorithm from @Salibian-Barrera01062006, which works well for small or moderate values of $p$. The algorithm works by taking a number of random subsamples of the data and computing S-estimates on these subsamples, and then picking the "best" candidate for the scale. It can be shown the number of subsamples required to return an S-estimator with $\epsilon$ breakdown point with probability $1-\alpha$ is 

$$N_{\text{sub}}\geq\frac{\log(\alpha)}{\log(1-(1-\epsilon)^p)}.$${#eq-subsample-lbound}

Taking $\alpha=0.01$, $\epsilon=0.5$ and $p=2$ for a slope and intercept term, the algorithm requires at least 17 subsamples of the data. @eq-subsample-lbound also shows that the number of subsamples scales exponentially in $p$. One potential concern with *Fast-S* is that it's non-deterministic and multiple re-runs will produce slightly different initial estimates. However, in practice the subsequent M-estimation still converges to a very similar final estimate in the end. Deterministic approaches such as @Peña01061999 can perform much better than *Fast-S* for large $p$, although in R libraries *Fast-S* is the more popular choice as an initial estimate for MM-estimation.

# Simulation study {#sec-simulation}

In this section we carry out simulations to compare the MM-estimator to ordinary least squares (OLS) on synthetic data. We aim to provide strong evidence that the proposed regression estimator has many desirable statistical properties. Simulations were carried out using the `furrr` package (@furrr), and regression lines were fit for $N=50, 200$ and $500$ with $10,000$ repetitions. Note that for results formatted as ratios, the OLS estimate is taken as the numerator. For simplicity, the following section deals with the estimate of the slope only.

## Normal efficiency simulation

Although MM-estimators were designed to deal with outliers and non-normal errors, it's also important to check they perform well in the case where the data has only irreducible noise. We take $y=x+\epsilon$ with $\epsilon\sim\mathcal{N}(0,1)$. The results can be found in @tbl-normal-eff. We see that the efficiency of 95% relative to OLS agrees with the value of $k=4.68$ used in the Bisquare M-estimator. Furthmore, the results show strong evidence that the M-estimator is consistent from the reduction of MSE as $N$ increases.

```{r}
#| label: tbl-normal-eff
#| tbl-cap: Comparison of variances of MM and OLS estimators with varying $N$, averaged over 10,000 repetitions. Efficiency is calculated as the ratio of variances, and note that as $\text{MSE}\approx\operatorname{Var}$ the efficiency and MSE ratio are identical after rounding.


mse <- function(x, x_hat) {
    mean((x - x_hat)^2)
}

# switching to a long format df can work too, but for only a few cols we prefer this
df <- read.csv("../data/normal_efficiency.csv")

df %>%
    group_by(N_pts) %>%
    summarise(
        OLS_mse = mse(slope_ols, 1),
        MM_mse = mse(slope_mm, 1),
        OLS_var = var(slope_ols),
        MM_var = var(slope_mm),
        efficiency = OLS_var / MM_var,
        mse_ratio = OLS_mse / MM_mse,
    ) %>%
    mutate(across(!c(efficiency, N_pts, mse_ratio), ~ formatC(.x, format = "e", digits = 3)), efficiency = round(efficiency, 2), mse_ratio = round(mse_ratio, 2)) %>%
    kbl(
        col.names = c(
            "$N$",
            "$\\text{MSE}(\\hat{\\beta}_{\\text{OLS}})$",
            "$\\text{MSE}(\\hat{\\beta}_{\\text{MM}})$",
            "$\\operatorname{Var}({\\hat{\\beta}_{\\text{OLS}}})$",
            "$\\operatorname{Var}({\\hat{\\beta}_{\\text{MSE}}})$",
            "Efficiency",
            "MSE ratio"
        ),
        format = "latex",
        booktabs = TRUE,
        escape = FALSE,
    ) %>%
    kable_styling(latex_options = c("hold_position"))
```

## Gross outlier contamination simulation

```{r}
#| label: tbl-gross-outlier-mse
#| tbl-cap: "Comparison of MSEs of MM and OLS estimators with varying $N$ and $\\mathcal{N}(0, 10^2)$ noise on 40% of observations, averaged over 10,000 repetitions. See the comment in @tbl-normal-eff regarding the efficiency and MSE ratio columns."

contamination_pct <- 40

df <- read.csv(glue("../data/gross_outlier_{contamination_pct}.csv"))

df %>%
    group_by(N_pts) %>%
    summarise(
        OLS_mse = mse(slope_ols, 1),
        MM_mse = mse(slope_mm, 1),
        OLS_var = var(slope_ols),
        MM_var = var(slope_mm),
        efficiency = OLS_var / MM_var,
        mse_ratio = OLS_mse / MM_mse,
    ) %>%
    mutate(across(!c(efficiency, N_pts, mse_ratio), ~ formatC(.x, format = "e", digits = 3)), efficiency = round(efficiency, 2), mse_ratio = round(mse_ratio, 2)) %>%
    kbl(
        col.names = c(
            "$N$",
            "$\\text{MSE}(\\hat{\\beta}_{\\text{OLS}})$",
            "$\\text{MSE}(\\hat{\\beta}_{\\text{MM}})$",
            "$\\operatorname{Var}({\\hat{\\beta}_{\\text{OLS}}})$",
            "$\\operatorname{Var}({\\hat{\\beta}_{\\text{MSE}}})$",
            "Efficiency",
            "MSE ratio"
        ),
        format = "latex",
        booktabs = TRUE,
        escape = FALSE,
    ) %>%
    kable_styling(latex_options = c("hold_position"))
```


# References 

::: {#refs}
:::


# Appendix {.appendix}

## Algorithms {#sec-algos}

```{=latex}
\newcommand{\Break}{\State \textbf{break}}
\algnotext{EndFor}
\algnotext{EndIf}
\algnotext{EndWhile}
\algnotext{EndFunction}

\begin{algorithm}[H]
\caption{Fast-S}\label{alg:fast-s}
\begin{algorithmic}
\Require $n \geq 0$
\Require $\text{length}(x) = \text{length}(y)$
\Require $T\gg0$ maximum number of iterations
\Function{Fast{-}S}{$x, y, p, T$}
    \State $\mathbf{X} \leftarrow \begin{bmatrix} \mathbf{1} & x \end{bmatrix}$ \Comment{add column of 1s for the intercept term}
    \State $\text{candidates} \leftarrow [\, ]$
    \State $N_{\text{sub}} \leftarrow \lceil \log(\alpha) / \log(1 - (1-\epsilon)^p) \rceil$ \Comment{see \eqref{eq-subsample-lbound}}
    \State  $\rho_k(x), w_k(x)$ \leftarrow \text{Bisquare $\rho$, and weight resp. with $k=1.547$}

    \For{$i = 1, \dots, N_{\text{sub}}$}
        \State $\xi_i \leftarrow $ random subsample of $p$ indices from $\{1, \dots, n\}$
        \State $\mathbf{X}_{i} \leftarrow X[\xi_i,\, :]$
        \State $y_{i} \leftarrow y[\xi_i]$
        \State $\hat{\beta} \leftarrow (\mathbf{X}_i^\top \mathbf{X}_i)^{-1} \mathbf{X}_i^\top y_i$ \Comment{fitting OLS through $p$ points}
        \State $r(\hat{\beta}) \leftarrow y - \mathbf{X} \hat{\beta}$ \Comment{take residuals on the full data}
        \State $s_{\text{init}} \leftarrow \Call{MADN}{r(\hat{\beta})}$
        \State $s \leftarrow \Call{S-scale-iteration}{r(\hat{\beta}), s_{\text{init}}, \, \rho_{k}(x)}$

        \State $K_{\text{iter}} \leftarrow 2$ \Comment{this can also be set to 1 with little effect on the output}
        \\
        
        \For{$j = 1, \dots, K_{\text{iter}}}$ \Comment{Do Improvement-steps (I-steps), similar to IRLS}

            \State $(\hat{\beta}, s ) \leftarrow \Call {I{-}Step}{\hat{\beta}, s, \, \rho_{k}(x), w_k(x)$}
        \EndFor
        \State $\text{candidates}[i] \leftarrow (\hat{\beta},\ s)$
    \EndFor

    \State $\text{candidates} \leftarrow$ the $\min(10, N_{\text{sub}})$ candidates with smallest $s$
    \\
    \State $s^* \leftarrow \infty$

    \For{$(\hat{\beta}, s) \in \text{candidates}$}
        \For{$i = 1 \dots T$} \Comment{I-steps until convergence}
            \State $(\hat{\beta}, s_{\text{new}}) \leftarrow \Call {I{-}Step}{\hat{\beta}, s, \, \rho_{k}(x), w_k(x)$}
            \If{$|(s_{\text{new}} / s) - 1| < \epsilon$} \Comment{convergence check}
                \State $s \leftarrow s_{\text{new}}$
                \Break
            \Else
                \State $s \leftarrow s_{\text{new}}$
            \EndIf
        \EndFor
        \If{$s < s^*$}
            \State $(\hat{\beta}^*, s^*) \leftarrow (\hat{\beta}, s)$
        \EndIf
    \EndFor

    \State \Return $(\hat{\beta}^*, s^*)$

    \EndFunction

\end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
\caption{S{-}Scale{-}iteration}\label{alg:s-scale}
\begin{algorithmic}
\Function{S{-}Scale{-}iteration}{$r(\hat{\beta}), s_{\text{init}}, \rho_k, w_k, T$}

    \State $b \leftarrow 0.5$ \Comment{chosen for the Bisquare family}

    \State $s \leftarrow s_{\text{init}}$
    \For{$k = 1, \dots, T$}
        \State $s_{\text{new}} \leftarrow s \sqrt{\frac{1}{nb} \sum_{i=1}^{n} \rho(r(\hat{\beta}) / s)}$ 
        \If{$|(s_{\text{new}} / s) - 1| < \epsilon$} \Comment{convergence check}
            \Break
        \EndIf
    \EndFor
    \State \Return $s$
\EndFunction

\end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
\caption{I{-}Step}\label{alg:i-step}
\begin{algorithmic}

\Function{I{-}Step}{$\hat{\beta}, s_{\text{init}}, \rho_k, w_k$}
    \State $\mathbf{W} \leftarrow \text{diag}(w_k(r(\hat{\beta}) / s))$
    \State $\hat{\beta} \leftarrow (\mathbf{X}^\top \mathbf{W} \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{W} y$
    \State $r(\hat{\beta}) \leftarrow y - \mathbf{X} \hat{\beta}$
    \State $s \leftarrow \Call{S-scale-iteration}{r(\hat{\beta}), s_{\text{init}}, \, \rho_{k}(x)}$
    \State \Return $(\hat{\beta}, s)$
\EndFunction

\end{algorithmic}
\end{algorithm}
```


## Normal distribution of MM-estimator slopes

First we check a histogram and QQplot of the MM-estimator slopes in @fig-slope-histogram-appendix. Visually it appears the slopes follow an approximately normal distribution. Then in @tbl-mm-ad-test-appendix we see the null hypothesis is rejected only for $N=50$, which can be attributed to it being a small sample.

```{r}
#| label: fig-slope-histogram-appendix
#| fig-cap: Histogram of slopes (above) and QQplot (below) of MM-estimator simulated in @sec-simulation for the $N=500$ case.
#| fig-width: 6
#| fig-height: 3


hist <- df %>%
    filter(N_pts == 500) %>%
    ggplot(aes(x = slope_mm)) +
    geom_histogram(bins = 30) +
    labs(x = TeX("$\\hat{\\beta}_{MM}$"), y = "Count") +
    theme_minimal()

qqplot <- df %>%
    filter(N_pts == 500) %>%
    ggplot(aes(sample = slope_mm)) +
    stat_qq(size = 0.15) +
    stat_qq_line() +
    theme_minimal()

hist / qqplot
```

```{r}
#| label: tbl-mm-ad-test-appendix
#| tbl-cap: Anderson-Darling (AD) test p-values by $N$.

df %>% group_by(N_pts) %>%
    summarise(ad_pval = ad.test(slope_mm)$p.value) %>%
    mutate(ad_pval = formatC(ad_pval, format = "e", digits = 3)) %>%
    kbl(
        col.names = c("$N$", "AD p-value"),
        format = "latex",
        booktabs = TRUE,
        escape = FALSE,
        align = "rc"
    ) %>%
    kable_styling(latex_options = c("hold_position"))
```