---
title: "Outlier Detection and Robust Linear Regression"
pdf-engine: xelatex
documentclass: article
date: today
fig-pos: "H"
geometry: margin=1in
execute:
    echo: false
    warning: false
    message: false
bibliography: references.bib
csl: cardiff-university-harvard.csl
link-citations: true
toc: true
toc-depth: 2
number-sections: true
fontsize: 11pt
author: 
    - "Adnan Akther (C23125434)"
    - "Ioan Barber (C23040485)"
    - "Anthony Lawrence (C23125134)"
    - "Llŷr Davies (C21077567)"
    - "Carwyn Hardiman (C23009849)"
    - "Ashley Woollard (C23065527)"
format:
    pdf:
        colorlinks: true
        toccolor: black
        linkcolor: blue
        citecolor: blue
        urlcolor: blue
        include-in-header:
            text: |
                \usepackage{amsmath}
                \usepackage{booktabs}
                \usepackage{float}
                \usepackage{unicode-math}
                \usepackage{algorithm}
                \usepackage{algpseudocode}
                \newcommand{\Break}{\State \textbf{break}}
abstract: |
    This report explores how traditional statistical models respond to outliers and assesses how well robust linear regression methods can handle them. Our main focus is on the MM-estimator, which is a method that uses a two-step approach combining the high breakdown point of an S-estimator with the high efficiency of an M-estimator. Simulations are conducted to verify statistical properties of the MM-estimator, which can be used as a robust alternative to OLS.
---

# Introduction

Outliers, for a long time, have been an important topic of discussion amongst data practitioners and statisticians - they constantly alter analysis and can cause undesirable or even false results. This is an extremely prevalent issue in areas that typically work on smaller data sets, such as clinical trials. For example, take @Cordero2025, which uses robust methods due to a small sample size. In this report we will discuss what is in outlier and some of the numerous methods to tackle them. Understanding and dealing with outliers first starts with rules of thumb, these approaches lack standardisation since they are created by individuals and are simply agreed upon as a viable approach by other data practitioners. This raises some clear issues, methods to avoid such lack of clarity have been developed for decades and many have had success. 
One of these methods is the M-estimator, this is a method that approximates the model of a data set with as little influence from outliers as possible. However, there are limitations with efficiency and a point where outliers become too much, even for such a method to handle. Another similar method is the Theil-Sen estimator (@Theil1992), it takes a different approach to avoid outliers and, when used alone, it is more resistent to outliers than the M-estimator. However, we found that the Theil-Sen estimator doesn’t scale well with multiple predictors nor does it have the same theoretical guarantees as other methods. To get started, in the first section we jump into simple methods which have been tried and tested whilst long awaiting improvement. For more discussion around Theil-Sen, see @sec-theil-sen.

# Simple Outlier Detection and Preliminaries

When collecting data, either a small or large amount, it is typically found that there are one or multiple ‘abnormal’ observations; we often call these observations outliers due to them not conforming to the majority of the data. However, it would be naïve to just call any observation that does not fall in line with trends to be an outlier or even worse to delete or manipulate it in order to not have to worry about the impact of this abnormal data on analysis methods. In this section, we discuss what should be required to confidently identify outliers in data, how these outliers can affect our statistical analysis, cover what options are available to deal with outliers and briefly cover some methods that can avoid outliers without us having to directly alter them.

If we consider the bulk of the data to be around the mean and have approximately symmetric distribution it can be observed that roughly 68%, 95%, and 99.7% of the data sits within one, two and three standard deviations of the mean respectively. The three sigma rule $(\mu \pm 3\sigma)$ is often cited as an indicator for an observation being an outlier, this rule of thumb utilises the standard deviation to detect observations outside of this 99.7% range; resulting in the data outside of this range either being changed to match the observation that is closest to $(\mu \pm 3\sigma)$, depending on which side of the interval the observation falls on, or to delete the observation from the sample. However the three-sigma rule has a few key flaws, one of which being that it is particularly prone to masking; masking can happen when multiple outliers are present causing each other to become undetected by this rule of thumb. A typical example of masking is having one or more extreme outliers which ‘mask’ the less extreme outliers, in this case the extreme outlier would drastically change the mean resulting in a shift of the $3\sigma$ rule bringing the less extreme outliers within the accepted range.

This issue can be avoided by other outlier detection methods, created with the intention of avoiding the problem of masking from extreme or clustered outliers. We define the Median Absolute Deviation (MAD) about the median in @eq-mad-formula as a robust alternative to the SD. Taking this with the median we can construct a robust $3\sigma$ rule, which involves replacing the sample mean with the median and the standard deviation with the MAD. 

$$
\text{MAD}(x) := \text{median}{|x_i - \text{median}(x_i)|}.
$$ {#eq-mad-formula}

However in this form the scale of the MAD is not comparable to the SD, so we divide by the MAD of a standard normal random variable, which we show in @sec-normal-dist-mad-proof to be $0.6745 = \Phi^{-1}(0.75)$. We denote the normalised MAD the MADN, therefore the robust $3\sigma$ rule can be written as 
$$
(\text{median}(x_i) \pm 3\cdot\text{MADN}(x_i)).
$$ {#eq-robust-3-sigma}


The drawbacks of using the median and MADN instead of the mean and standard deviation are found a loss of statistical efficiency. The inefficiency of MAD becomes an issue when dealing with data that has no outliers and we therefore are reducing efficiency for little to no gain in accuracy or precision. However, there are other more popular outlier detection methods.

## OLS for outlier detection and Cook's distance

The final method we will discuss is utilising OLS (Ordinary Least Squares) to uncover influential observations, we will be talking about bivariate data from here on. This is done by finding the line that best fits the data and then seeing which observations have high leverage and large residual. A point of high leverage is one that sits near an edge of the majority of data, meaning we are only concerned for the x value when looking at leverage. To understand when to call a point influential or not we must also discuss residuals of observations, the residual of an observation is simply the distance of the y value, given by the observation, from the predicted value of the fitted model. When an observation has enough leverage and a high residual it is then called influential. The ultimate decision of an observation being significantly influential is still left up to the data practitioner, @Cook_1977 came up with a way to combine the leverage with the residual of an observation in attempt to quantify how influential any observation is. The Cook's distance formula is shown in @eq-cooks-distance-formula, with $h_ii$ being diagonal elements of the "hat matrix"^[Sometimes the $h_{ii}$ are called leverages]. When used, Cook's distance is alongside a chosen rule of thumb to characterise a point as influential, one rule of thumb is comparing the Cook’s distance of observations to the F distribution with $D_i > F_{p, n-p}(0.5)$ means that observation is highly influential, even values of $D_i > F_{p, n-p}(0.2)$ are typically investigated. 
$$ 
\frac{1}{p+1}\left(\frac{h_{ii}}{1-{h_{ii}}}\right)r_i^2
$$ {#eq-cooks-distance-formula}

Another such rule of thumb is if $D_i > (4/(n - p - 1))$ then the observation is of significant influence, with n being the number of points and p being the number of observations. Similarly to before, what follows is a decision on whether the observation should be kept as is, brought within an 'acceptable' range or deleted from the data sample/set. However, there is debate whether Cook's distance accurately represents outliers, for example in @Kim2017.

A recurring issue with these outlier detection methods that we have covered is that they require human input on what results flag an observation as an outlier. This lack of systemic practice creates valid scrutiny for these methods, and as such we want a way to avoid the impact of these outliers without human intervention and taking assumptions which require in-depth understanding of the data collection. One strong set of methods with extensive theoretical backing is M-estimation @alma9911772489302420[p. 22]
which generalises OLS and allows for control, via a parameter, over the robustness of the estimator. However, increasing the robustness has a trade-off of decreasing the efficiency under normal errors. Including M-estimators, regression estimators have a property called breakdown points detailing what the largest proportion of contamination the data can have before guaranteeing the estimation to be totally uninformative. Intuitively, the breakdown point of an estimator can't be higher than 50% - if more than half of data is noise then the estimator will be unable to create an uncontaminated model. In the next section, we will start with M-estimators to explore novel methods for data analysis of contaminated data whilst considering the cost of efficiency if the data were clean. For discussion of other methods to detect outliers, see `EllipticEnvelope` from @scikit-learn and @alma9911772489302420[p.207-210]

# M-Estimators {#sec-m-estimators}

Many models applying statistics or classification are highly sensitive to outliers. Even one extreme outlier can skew estimates, resulting in unreliable results. Therefore, we must ensure stable estimation methods to manoeuvre around these problematic outliers.

One of the main estimators in robust statistics is the M-estimator, meaning maximum likelihood–type estimator. These provide a flexible method for estimating parameters.

They are defined as solutions to optimisation problems by choosing a loss function in @alma9911772489302420[p. 23-25]. Let $x_1, \dots, x_n$ be observations and let $\rho(\cdot)$ be a loss function. An M-estimator of parameter $\theta$ is defined as

$$
\hat{\theta} = \arg\min_\theta \sum_{i=1}^n \rho(x_i - \theta).
$$

We obtain the estimating equations by differentiating the objective function. Let

$$
\psi(u) = \rho'(u).
$$

Then the estimator satisfies

$$
\sum_{i=1}^n \psi(x_i - \hat{\theta}) = 0.
$$

Different choices of $\rho$ lead to different estimators. For example, absolute loss yields the sample median, while squared loss returns the sample mean.

M-estimators can be used for an array of parameters including location, scale, and regression coefficients. Location M-estimators provide alternatives to the sample mean, scale estimators provide robustness in measuring dispersion, and in this report we focus on regression M-estimators, which can be interpreted as weighted least squares.

## Regression M-Estimators

For robust regression we consider the linear regression model

$$
y_i = x_i^T \beta + \varepsilon_i, \quad i = 1, \dots, n
$$

where $x_i$ is a vector of predictors, $\beta$ is the vector of regression coefficients and $\varepsilon_i$ is a irreducible noise vector. The residual corresponding to a parameter value $\beta$ is

$$
r_i(\beta) = y_i - x_i^T \beta.
$$

Regression M-estimators are defined in @alma9911772489302420[p. 98] by minimizing a robust objective function of the standardised residuals

$$
\hat{\beta} =
\arg\min_\beta
\sum_{i=1}^n
\rho \left( \frac{r_i(\beta)}{s} \right),
$$ {#eq-m-estimator-minimiser}

where $s$ is a scale estimate used to standardise the residuals.

The corresponding estimating equations are

$$
\sum_{i=1}^n
\psi \left( \frac{r_i(\beta)}{s} \right) x_i = 0,
\quad \text{where} \quad \psi = \rho'.
$$

These equations can also be written in the form

$$
\sum_{i=1}^n w_i r_i(\beta)x_i = 0 \quad \text{with weights} \quad 
w_i = \frac{\psi\left(\frac{r_i(\beta)}{s}\right)}{\frac{r_i(\beta)}{s}}.
$$


Since many quantities depend on one another, the estimation procedure is iterative. The residuals depend on the estimate of $\beta$, the weights $w_i$ depend on the residuals through the $\psi$ function, and the updated estimate of $\beta$ depends on the weights.

Typically, the estimation is performed using iteratively reweighted least squares (IRLS), see the I-step algorithm in @sec-algos for more information. We start from an initial estimate of $\beta$, compute the residuals, determine the weights from the $\psi$ function, and then compute a weighted least squares fit. This procedure is repeated until convergence. More discussion can be found in @alma9911772489302420[p. 104].

A scale estimate $s$ is used to standardise the objective function. The scale can either be fixed or estimated jointly with the regression parameters. It is important that $s$ is robust, since a non-robust scale could become very large when outliers are present. In that case the standardised residuals $\frac{r_i(\beta)}{s}$ become artificially small and the weighting mechanism no longer reduces the influence of extreme observations.

A common choice for the initial scale estimate is the normalized median absolute deviation (MADN).

The behaviour of the estimator is determined by the choice of loss function $\rho$, its derivative $\psi$, and the associated weight function $W$.

```{r}
#| label: library-setup
library(knitr)
library(kableExtra)
library(tidyverse)
library(glue)
library(latex2exp)
library(patchwork)
library(nortest)

set.seed(0)

source("../functions.R")
```
```{r}
#| label: fig-psi-rho-weight-functions
#| fig-cap: Plots of $\rho$, $\psi$ and weight functions of Huber and Bisquare M-estimators. See @eq-huber-m-estimator and @eq-bisquare-m-estimator for more information.
#| fig-width: 10
#| fig-height: 6

k <- 1.548
x_vals <- seq(-3, 3, length.out = 100)
plotting_df <- data.frame(
    x = x_vals, h_rho = huber_rho(x_vals, k),
    h_psi = huber_psi(x_vals, k = k),
    bi_rho = bisquare_rho(x_vals, k = k),
    bi_psi = bisquare_psi(x_vals, k = k),
    h_weight = huber_weight(x_vals, k = k),
    bi_weight = bisquare_weight(x_vals, k = k)
)
# maybe the following can be condensed...

p1 <- ggplot(plotting_df) +
    geom_line(aes(x = x, y = h_rho), color = "blue") +
    geom_line(aes(x = x, y = bi_rho), color = "red") +
    labs(x = TeX("$x$"), y = TeX("$\\rho(x)$")) +
    theme_minimal()

p2 <- ggplot(plotting_df) +
    geom_line(aes(x = x, y = h_psi, color = "Huber")) +
    geom_line(aes(x = x, y = bi_psi, color = "Bisquare")) +
    scale_color_manual(values = c("Huber" = "blue", "Bisquare" = "red")) +
    labs(x = TeX("$x$"), y = TeX("$\\psi(x)$"), color = "Estimator") +
    theme_minimal()

p3 <- ggplot(plotting_df) +
    geom_line(aes(x = x, y = h_weight), color = "blue") +
    geom_line(aes(x = x, y = bi_weight), color = "red") +
    labs(x = TeX("$x$"), y = "Weight") +
    theme_minimal()

p1 / p2 / p3 # this is from patchwork
```


Two common choices are the Huber and Bisquare functions given by @eq-huber-m-estimator, and @eq-bisquare-m-estimator respectively. The functions are plotted in @fig-psi-rho-weight-functions. The Huber function limits the influence of large residuals but never assigns a weight of zero, meaning extreme observations still have a small influence on the fit. In contrast, the Bisquare function assigns zero weight to sufficiently large residuals, effectively removing these observations from the fitting procedure. M-estimators which satisfy this are called *redescending* and they are the most popular for linear regression, despite them not necessarily producing a unique solution of @eq-m-estimator-minimiser. 

Varying the value of $k$ in @eq-huber-m-estimator and @eq-bisquare-m-estimator controls the point at which residuals are down-weighted, and it can be shown increasing $k$ increases the efficiency (under normal errors) of the M-estimator. However, this increase causes fewer outliers to be down-weighted, reducing robustness. Taking $k\to\infty$ in these equations the M-estimators converge to OLS, which can also be seen through the weight functions in @fig-psi-rho-weight-functions. In practice, @alma9911772489302420 recommends choosing $k$ so that the estimator is 85 or 95% efficient under normal errors.
$$
\text{Huber:}\, \rho_k=
\begin{cases} 
    x^2 & |x|\leq k \\
    2k|x| - k^2 & |x|>k
\end{cases}
\qquad
\psi_k(x) =
\begin{cases} 
    2x & |x|\leq k \\
    2k\operatorname{sgn}(x) & |x|>k
\end{cases}
$$ {#eq-huber-m-estimator}

$$
\text{Bisquare:}\, \rho_k=
\begin{cases} 
    1 - [1 - (x/k)^2]^3 & |x|\leq k \\
    1 & |x|>k
\end{cases}
\qquad
\psi_k(x) = \frac{6}{k^2}x\left[1-\left(\frac{x}{k}\right)^2\right]\mathbb{1}_{|x|\leq k }
$$ {#eq-bisquare-m-estimator}


$$
\text{with weight function} \quad W_k(x) = \frac{\psi_k(x)}{x}
$$

While regression M-estimators improve robustness compared to ordinary least squares, the initial fit can still be sensitive to outliers. Even with a robust initial scale, large outliers may influence the early iterations of the algorithm. Consequently, regression M-estimators may not be sufficiently robust in all situations. This motivates the development of more robust procedures such as S-estimators, which are discussed in the following section.

## Regression S-estimators

S-estimators are instead defined as minimisers of the scale
$$\hat{\beta} = \operatorname*{argmin}_{\beta} s(r(\beta)),$$

where $s(r)$ is taken to be an M-estimator of scale
$$
\frac{1}{n}\sum_{i=1}^{n}{\rho\left(\frac{r_i(\beta)}{s}\right)} = \delta,
$$ {#eq-implicit-m-scale}

and other quantities defined as in @sec-m-estimators. Although the formula is very similar to @eq-m-estimator-minimiser, it is not an M-estimator as they solve a different optimisation problem. This is because @eq-implicit-m-scale is defined implicitly as changes to $\beta$ also affect $s(r(\beta))$. S-estimators with bounded $\rho$ functions have the property that their breakdown point can be set to 50%, the maximum possible value. It can be shown the constant $\delta$ controls the breakdown point of the estimator, which is given by 

$$ \epsilon = \min(1-\delta, \delta) $$

which has a maximum at $\delta = 0.5$. Once this is chosen, it must satisfy the normalisation condition
$$\mathbb{E}\left[\rho_k\left(\frac{x}{s}\right)\right]=\delta \quad \text{with} \quad x\sim\mathcal{N}(0,1)$$

so that under standard normal errors we get a scale estimate of 1. More discussion can be found in @alma9911772489302420[p. 34, 60] This constraint also determines $k$ for the S-estimator, which sets the efficiency^[For the Bisquare M-estimator defined in @sec-m-estimators, choosing $\delta=0.5$ here leads to $k=1.56$]. However despite this high breakdown point, it can be shown that S-estimators with smooth $\rho$ under normal errors can't have an efficiency higher than 0.33, see @alma9911772489302420[p. 125-126] for more information. In @sec-mm-and-fast-s we will resolve this with a procedure that is able to combine the high breakdown point of the S-estimator with the high efficiency of an M-estimator under normal errors.

# Regression MM-Estimators and Fast-S {#sec-mm-and-fast-s}

MM-estimators were introduced by @10.1214/aos/1176350366 and work on a two-step process of starting with an S-estimator to robustly estimate scale, followed by M-estimation with fixed scale coming from the previous step. The S-estimator also provides a starting point for estimating $\beta$ in the second step. MM-estimators are able to inherit the high breakdown point of the S-estimator^[provided $k$>$k_0$, where k is the $k$ parameter from the regression M-estimator, and $k_0$ is from the S-estimator], while also keeping the high normal efficiency of an M-estimator. The M-estimation step follows @eq-m-estimator-minimiser as in the previous section.

Let $p$ denote the number of predictors in a regression problem. Computing an S-estimator is computationally challenging, but there are viable approximations for $p$ small. We have implemented the *Fast-S* algorithm from @Salibian-Barrera01062006, which works well for small or moderate values of $p$. The algorithm works by taking a number of random subsamples of the data and computing S-estimates on these subsamples, and then picking the "best" candidate for the scale. It can be shown the number of subsamples required to return an S-estimator with $\epsilon$ breakdown point with probability $1-\alpha$ is 

$$N_{\text{sub}}\geq\frac{\log(\alpha)}{\log(1-(1-\epsilon)^p)}. $$ {#eq-subsample-lbound}

Taking $\alpha=0.01$, $\epsilon=0.5$ and $p=2$ for a slope and intercept term, the algorithm requires at least 17 subsamples of the data. @eq-subsample-lbound also shows that the number of subsamples scales exponentially in $p$. For the full pseudocode, see @sec-algos. One potential concern with *Fast-S* is that it's non-deterministic and multiple re-runs will produce slightly different initial estimates. However, in practice the subsequent M-estimation still converges to a very similar final estimate in the end. Deterministic approaches such as @Peña01061999 (P-Y) can perform much better than *Fast-S* for large $p$, although in R libraries *Fast-S* is the more popular choice as an initial estimate for MM-estimation.

# Simulation study {#sec-simulation}

In this section we carry out simulations to compare the MM-estimator to ordinary least squares (OLS) on synthetic data. We aim to provide strong evidence that the proposed regression estimator has many desirable statistical properties. Simulations were carried out using the `furrr` package (@furrr), and regression lines were fit for $N=50, 200$ and $500$ with $10,000$ repetitions. Simulations start with data from 

$$ y=x+\epsilon \quad \text{where} \quad \epsilon\sim\mathcal{N}(0,1) $$ {#eq-line-and-small-gaussian-noise}

and add additional noise as needed. Note that for results formatted as ratios, the OLS estimate is taken as the numerator. Unless stated otherwise, the following sections mostly deal with the estimate of the slope and discussion of the intercept term can be found in the appendix.

## Normal efficiency simulation {#sec-normal-eff-sim}

Although MM-estimators were designed to deal with outliers and non-normal errors, it's also important to check they perform well in the case where the data has only irreducible normal noise. For this we take samples from @eq-line-and-small-gaussian-noise only. The results can be found in @tbl-normal-eff. We see that the efficiency of 95% relative to OLS agrees with the value of $k=4.68$ used in the Bisquare M-estimator. We also verify the normal distribution of the slopes in @sec-m-est-normal-dist. Furthmore, the results show strong evidence that the M-estimator is consistent from the reduction of MSE as $N$ increases. We deal with adding noise to the observations in the following subsections.

```{r}
#| label: tbl-normal-eff
#| tbl-cap: Comparison of variances of MM and OLS estimators with varying $N$, averaged over 10,000 repetitions. Efficiency is calculated as the ratio of variances, and note that as $\text{MSE}\approx\operatorname{Var}$ the efficiency and MSE ratio are identical after rounding.


mse <- function(x, x_hat) {
    mean((x - x_hat)^2)
}

rmse <- function(x, x_hat) {
    sqrt(mse(x, x_hat))
}

# switching to a long format df can work too, but for only a few cols we prefer this
df <- read.csv("../data/normal_efficiency.csv")

df %>%
    group_by(N_pts) %>%
    summarise(
        OLS_mse = mse(slope_ols, 1),
        MM_mse = mse(slope_mm, 1),
        OLS_var = var(slope_ols),
        MM_var = var(slope_mm),
        efficiency = OLS_var / MM_var,
        mse_ratio = OLS_mse / MM_mse,
    ) %>%
    mutate(across(!c(efficiency, N_pts, mse_ratio), ~ formatC(.x, format = "e", digits = 3)), efficiency = round(efficiency, 2), mse_ratio = round(mse_ratio, 2)) %>%
    kbl(
        col.names = c(
            "$N$",
            "$\\text{MSE}(\\hat{\\beta}_{\\text{OLS}})$",
            "$\\text{MSE}(\\hat{\\beta}_{\\text{MM}})$",
            "$\\operatorname{Var}({\\hat{\\beta}_{\\text{OLS}}})$",
            "$\\operatorname{Var}({\\hat{\\beta}_{\\text{MM}}})$",
            "Efficiency",
            "MSE ratio"
        ),
        format = "latex",
        booktabs = TRUE,
        escape = FALSE,
    ) %>%
    kable_styling(latex_options = c("hold_position"))
```

## Gross outlier contamination simulation {#sec-g-outlier-sim}

Next we simulate the MM-estimator again with more more realistic noise. Starting with @eq-line-and-small-gaussian-noise again, 40% of the data points have $\mathcal{N}(0, 10)$ noise added. MM and OLS estimates are fitted and results are shown in @tbl-gross-outlier-mse. In this situation we see that the MM-estimator is an order of magnitude more efficient and less variable compared to OLS, although OLS does not fully breakdown.

```{r}
#| label: tbl-gross-outlier-mse
#| tbl-cap: "Comparison of MSEs of MM and OLS estimators with varying $N$ and $\\mathcal{N}(0, 10^2)$ noise on 40% of observations, averaged over 10,000 repetitions. See the comment in @tbl-normal-eff regarding the efficiency and MSE ratio columns."

contamination_pct <- 40

df <- read.csv(glue("../data/gross_outlier_{contamination_pct}.csv"))

df %>%
    group_by(N_pts) %>%
    summarise(
        OLS_mse = mse(slope_ols, 1),
        MM_mse = mse(slope_mm, 1),
        OLS_var = var(slope_ols),
        MM_var = var(slope_mm),
        efficiency = OLS_var / MM_var,
        mse_ratio = OLS_mse / MM_mse,
    ) %>%
    mutate(across(!c(efficiency, N_pts, mse_ratio), ~ formatC(.x, format = "e", digits = 3)), efficiency = round(efficiency, 2), mse_ratio = round(mse_ratio, 2)) %>%
    kbl(
        col.names = c(
            "$N$",
            "$\\text{MSE}(\\hat{\\beta}_{\\text{OLS}})$",
            "$\\text{MSE}(\\hat{\\beta}_{\\text{MM}})$",
            "$\\operatorname{Var}({\\hat{\\beta}_{\\text{OLS}}})$",
            "$\\operatorname{Var}({\\hat{\\beta}_{\\text{MM}}})$",
            "Efficiency",
            "MSE ratio"
        ),
        format = "latex",
        booktabs = TRUE,
        escape = FALSE,
    ) %>%
    kable_styling(latex_options = c("hold_position"))
```

## Breakdown point simulation {#sec-breakdown-sim}

Now we seek to verify the breakdown point of the MM-estimator constructed in @sec-mm-and-fast-s, which is roughly 50%. In this simulation we start from @eq-line-and-small-gaussian-noise once again and we set an increasing proportion of points to $y=10,000$. This sets up two competing lines, namely $y = 10000$ and @eq-line-and-small-gaussian-noise. We fix $N=200$ and $1500$ replications at contamination proportions of $\epsilon=0.2, 0.45, 0.49, 0.5, 0.51$ and $0.55$. The results can be found in @fig-breakdown-line-graph and @tbl-breakdown-slope, along with discussion of the intercept term in @tbl-breakdown-intercept-appendix. The most striking feature of the figure is the spike near $\epsilon=0.5$ and the RMSE then decreasing back to one - which is actually the error when the model fits only to the $y=10000$ line. The highest RMSE represents the point when the MM-estimator can't distinguish the contamination from the original line, but for $\epsilon$ bigger than 0.5 the $y=10000$ the MM-estimator exclusively passes through contaminated points. A similar story is true for the intercept of the MM-estimator (lower panel), that it degrades slowly until half the points are contaminated. On the other hand, in both cases OLS performs badly even under $\epsilon=0.2$, and it degrades at a constant rate. In @tbl-breakdown-slope we see that the variance also behaves similar to the RMSE for the MM-estimator, being highest when noise and real data have equal proportions and then decreasing when noise is more frequent than real data. An interesting finding is that this does not occur for the bias of the slope, however for this type of noise a bias near 1 still shows breakdown of the estimator.


```{r}
#| label: fig-breakdown-line-graph
#| fig-cap: Line graph of RMSE (y) against $\epsilon$ proportion of outliers. MSE could be used instead for the y-axis, but RMSE makes for easier interpretation. Note the gradual worsening of OLS and the sudden increase of RMSE for MM near $\epsilon=0.5$, verifying the breakdown point. Shaded bootstrap 95% confidence intervals are plotted also, calculated over 1000 samples with replacement.
#| fig-width: 6
#| fig-height: 4

df <- read.csv("../data/breakdown_200.csv")

boot_rmse <- function(x, x_hat, reps = 1000) {
    boots <- replicate(reps, {
        s <- sample(x, replace = TRUE)
        sqrt(mean((s - x_hat)^2))
    })
    quantile(boots, c(0.025, 0.975))
}

# long code blocks here, would need to be refactored if we were considering more predictors or different estimators
slope_df <- df %>%
    group_by(cp) %>%
    summarise(
        rmse_mm = sqrt(mean((slope_mm - 1)^2)),
        rmse_ols = sqrt(mean((slope_ols - 1)^2)),
        mm_ci_lo = boot_rmse(slope_mm, 1)[1],
        mm_ci_hi = boot_rmse(slope_mm, 1)[2],
        ols_ci_lo = boot_rmse(slope_ols, 1)[1],
        ols_ci_hi = boot_rmse(slope_ols, 1)[2]
    )

intercept_df <- df %>% # the true value of our intercept
    group_by(cp) %>%
    summarise(
        rmse_mm = sqrt(mean((intercept_mm - 0)^2)),
        rmse_ols = sqrt(mean((intercept_ols - 0)^2)),
        mm_ci_lo = boot_rmse(intercept_mm, 0)[1],
        mm_ci_hi = boot_rmse(intercept_mm, 0)[2],
        ols_ci_lo = boot_rmse(intercept_ols, 0)[1],
        ols_ci_hi = boot_rmse(intercept_ols, 0)[2]
    )

# ---- 

slope_breakdown <- ggplot(slope_df, aes(x = cp)) +
    geom_ribbon(aes(ymin = mm_ci_lo, ymax = mm_ci_hi), fill = "red", alpha = 0.2) +
    geom_ribbon(aes(ymin = ols_ci_lo, ymax = ols_ci_hi), fill = "blue", alpha = 0.2) +
    geom_line(aes(y = rmse_mm, color = "MM")) +
    geom_line(aes(y = rmse_ols, color = "OLS")) +
    scale_color_manual(values = c("MM" = "red", "OLS" = "blue"), name = NULL) +
    labs(x = NULL, title = "Slope", y = "RMSE") +
    theme_minimal() +
    theme(axis.text.x = element_blank(), legend.position = "none", plot.title = element_text(size = 9))

intercept_breakdown <- ggplot(intercept_df, aes(x = cp)) +
    geom_ribbon(aes(ymin = mm_ci_lo, ymax = mm_ci_hi), fill = "red", alpha = 0.2) +
    geom_ribbon(aes(ymin = ols_ci_lo, ymax = ols_ci_hi), fill = "blue", alpha = 0.2) +
    geom_line(aes(y = rmse_mm, color = "MM")) +
    geom_line(aes(y = rmse_ols, color = "OLS")) +
    scale_color_manual(values = c("MM" = "red", "OLS" = "blue"), name = NULL) +
    labs(title = "Intercept", x = TeX("$\\epsilon$"), y = "RMSE") +
    theme_minimal() +
    theme(legend.position = "none", plot.title = element_text(size = 9))

(slope_breakdown / intercept_breakdown +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")) +
    plot_annotation(title = TeX("RMSE breakdown of $\\hat{\\beta}$"), theme = theme(plot.title = element_text(size = 11)))

# ----
```

```{r}
#| label: tbl-breakdown-slope
#| tbl-cap: Table of RMSE, variance and bias of OLS and MM estimators for varying proportion $\epsilon$ of contamination. Contaminated values are set to 10000, whereas clean data follows@eq-line-and-small-gaussian-noise. This table deals with the slope estimate, see @sec-breakdown-intercept-appendix for discussion of the intercept. 

df %>%
    group_by(cp) %>%
    summarise(
        OLS_rmse = rmse(slope_ols, 1),
        MM_rmse = rmse(slope_mm, 1),
        OLS_var = var(slope_ols),
        MM_var = var(slope_mm),
        OLS_bias = mean(slope_ols) - 1,
        MM_bias = mean(slope_mm) - 1
    ) %>%
    mutate() %>%
    kbl(
        col.names = c(
            "$\\epsilon$",
            "$\\text{RMSE}(\\hat{\\beta}_{\\text{OLS}})$",
            "$\\text{RMSE}(\\hat{\\beta}_{\\text{MM}})$",
            "$\\operatorname{Var}({\\hat{\\beta}_{\\text{OLS}}})$",
            "$\\operatorname{Var}({\\hat{\\beta}_{\\text{MM}}})$",
            "$\\text{Bias}({\\hat{\\beta}_{\\text{OLS}}})$",
            "$\\text{Bias}({\\hat{\\beta}_{\\text{MM}}})$"
        ),
        format = "latex",
        booktabs = TRUE,
        escape = FALSE,
    ) %>%
    kable_styling(latex_options = c("hold_position"))
```

## Agreement with MASS

We lastly verify our implementation is correct by testing against `MASS` @MASS, specifically the `rlm` function which performs MM-estimation with Fast-S as the initial estimator. 

```{r}
#| label: fig-mass-vs-mm-convergence
#| fig-cap: Line graph of number of samples (x) against relative and absolute difference between our MM-estimator and MASS, for the slope and intercept respectively. Note the absolute difference was chosen for the intercept to avoid division by zero. The median and 95th percentiles are shown, and the decrease as N increases shows our estimator agrees very closely with MASS.
#| fig-width: 6
#| fig-height: 4

df <- read.csv("../data/mass_agreement.csv")
plot_df <- df %>%
    mutate(rd_slope = abs(slope_mm - slope_mass), ad_int = abs(intercept_mm - intercept_mass)) %>%
    group_by(N_pts) %>%
    summarise(
        median_rd_slope = median(rd_slope),
        median_ad_int = median(ad_int),
        pct_95_rd_slope = quantile(rd_slope, 0.95),
        pct_95_ad_int = quantile(ad_int, 0.95)
    )

# thought about adding a bootstrap ci but it's so small it's invisible!

p1 <- ggplot(plot_df, aes(x = N_pts)) +
    geom_line(aes(y = median_rd_slope, color = "median")) +
    geom_line(aes(y = pct_95_rd_slope, color = "95th percentile"), linetype = 2) +
    scale_colour_manual(values = c("median" = "red", "95th percentile" = "blue"), name = NULL) +
    labs(x = NULL, y = "relative", title = "Slope") +
    theme_minimal() +
    theme(axis.text.x = element_blank(), legend.position = "none", plot.title = element_text(size = 9))

p2 <- ggplot(plot_df, aes(x = N_pts)) +
    geom_line(aes(y = median_ad_int, color = "median")) +
    geom_line(aes(y = pct_95_ad_int, color = "95th percentile"), linetype = 2) +
    scale_colour_manual(values = c("median" = "red", "95th percentile" = "blue"), name = NULL) +
    labs(x = TeX("$N$"), y = "absolute", title = "Intercept") +
    theme_minimal() +
    theme(legend.position = "none", plot.title = element_text(size = 9))


(p1 / p2 + 
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")) +
    plot_annotation(title = "Agreement between MASS and our MM-estimator by sample size", theme = theme(plot.title = element_text(size = 11)))
```

# Conclusion


To conclude, by motivating examples from rules-of-thumb and heuristic approaches for outlier detection we introduce M-estimators as a generalisation of MLE and OLS. Next, the theory S and MM-estimators and their robust properties is discussed. This represents a systematic approach to down-weighting outliers which is preferable to removal of data points and other methods without having assumptions about the data. MM-estimators retain many statistical properties of OLS, including consistency and asymptotic normality, while having a breakdown point of 50% and remaining 85-95% efficient relative to OLS under normal errors. Practically, the Fast-S algorithm allows for MM-estimators to be fitted quickly for moderate $p$, and for large $p$ the P-Y initial estimator from @sec-mm-and-fast-s performs very well. Simulations in @sec-simulation verify this further - MM-estimation is a viable robust alternative to OLS. 

The procedure described in @sec-mm-and-fast-s can be used with different $\psi$ functions, and as an initial estimator to further increase efficiency and finite-sample breakdown point, see @alma9911772489302420[p. 155, 159-164]. In higher dimensions, outliers are harder to visualise and account for directly, emphasising the importance of MM-estimators for robust multivariate regression. This is not limited to simple regresion problems, for instance regularisation and time-series analysis are also amenable to robust methods, see @alma9911772489302420[p. 165-171, 293-361].



# References 

::: {#refs}
:::

# Author Contributions

- Conceptualisation: AL, AW, LlD
- Project administration: AL
- Software: AL, AW, IB
- Formal Analysis: AL, IB
- Validation: AL
- Visualisation: AL, LlD
- Writing - original draft: AL, AW, LlD, CH, IB
- Writing - review and editing: AL, AW, LlD, CH, IB, AA
- Welsh translation: CH, LlD, IB

# Appendix {.appendix}

## Algorithms {#sec-algos}

\begin{algorithm}[H]
\begin{algorithmic}
\caption{Fast-S}
\label{alg:fast-s}
\algnotext{EndFor}
\algnotext{EndIf}
\algnotext{EndWhile}
\algnotext{EndFunction}

\Require $n \geq 0$
\Require $\text{length}(x) = \text{length}(y)$
\Require $T\gg0$ maximum number of iterations
\Function{Fast{-}S}{$x, y, p, T$}
    \State $\mathbf{X} \leftarrow \begin{bmatrix} \mathbf{1} & x \end{bmatrix}$ \Comment{add column of 1s for the intercept term}
    \State $\text{candidates} \leftarrow [\, ]$
    \State $N_{\text{sub}} \leftarrow \lceil \log(\alpha) / \log(1 - (1-\epsilon)^p) \rceil$ \Comment{see \eqref{eq-subsample-lbound}}
    \State  $\rho_k(x), w_k(x)$ \leftarrow \text{Bisquare $\rho$, and weight resp. with $k=1.547$}


    \For{$i = 1, \dots, N_{\text{sub}}$}
        \State $\xi_i \leftarrow $ random subsample of $p$ indices from $\{1, \dots, n\}$
        \State $\mathbf{X}_{i} \leftarrow \mathbf{X}[\xi_i,\, :]$
        \State $y_{i} \leftarrow y[\xi_i]$
        \State $\hat{\beta} \leftarrow (\mathbf{X}_i^\top \mathbf{X}_i)^{-1} \mathbf{X}_i^\top y_i$ \Comment{fitting OLS through $p$ points}
        \State $r(\hat{\beta}) \leftarrow y - \mathbf{X} \hat{\beta}$ \Comment{take residuals on the full data}
        \State $s_{\text{init}} \leftarrow \Call{MADN}{r(\hat{\beta})}$
        \State $s \leftarrow \Call{S-scale-iteration}{r(\hat{\beta}), s_{\text{init}}, \, \rho_{k}(x)}$

        \State $K_{\text{iter}} \leftarrow 2$ \Comment{this can also be set to 1 with little effect on the output}
        \\
        
        \For{$j = 1, \dots, K_{\text{iter}}}$ \Comment{Do Improvement-steps (I-steps), similar to IRLS}

            \State $(\hat{\beta}, s ) \leftarrow \Call {I{-}Step}{\hat{\beta}, s, \, \rho_{k}(x), w_k(x)$}
        \EndFor
        \State $\text{candidates}[i] \leftarrow (\hat{\beta},\ s)$
    \EndFor

    \State $\text{candidates} \leftarrow$ the $\min(10, N_{\text{sub}})$ candidates with smallest $s$
    \\
    \State $s^* \leftarrow \infty$

    \For{$(\hat{\beta}, s) \in \text{candidates}$}
        \For{$i = 1 \dots T$} \Comment{I-steps until convergence}
            \State $(\hat{\beta}, s_{\text{new}}) \leftarrow \Call {I{-}Step}{\hat{\beta}, s, \, \rho_{k}(x), w_k(x)$}
            \If{$|(s_{\text{new}} / s) - 1| < \epsilon$} \Comment{convergence check}
                \State $s \leftarrow s_{\text{new}}$
                \Break
            \Else
                \State $s \leftarrow s_{\text{new}}$
            \EndIf
        \EndFor
        \If{$s < s^*$}
            \State $(\hat{\beta}^*, s^*) \leftarrow (\hat{\beta}, s)$
        \EndIf
    \EndFor

    \State \Return $(\hat{\beta}^*, s^*)$

    \EndFunction

\end{algorithmic}
\end{algorithm}


\begin{algorithm}[H]
\begin{algorithmic}
\caption{S{-}Scale{-}iteration}
\label{alg:s-scale-iteration}
\Function{S{-}Scale{-}iteration}{$r(\hat{\beta}), s_{\text{init}}, \rho_k, w_k, T$}

    \State $\delta \leftarrow 0.5$ \Comment{chosen for the Bisquare family}

    \State $s \leftarrow s_{\text{init}}$
    \For{$k = 1, \dots, T$}
        \State $s_{\text{new}} \leftarrow s \sqrt{\frac{1}{n\delta} \sum_{i=1}^{n} \rho(r(\hat{\beta}) / s)}$ 
        \If{$|(s_{\text{new}} / s) - 1| < \epsilon$} \Comment{convergence check}
            \State $s \leftarrow s_{\text{new}}$
            \Break
        \Else
            \State $s \leftarrow s_{\text{new}}$
        \EndIf
    \EndFor
    \State \Return $s$
\EndFunction

\end{algorithmic}
\end{algorithm}

\begin{algorithm}[H]
\begin{algorithmic}
\caption{I{-}Step}
\label{alg:I-step}

\Function{I{-}Step}{$\hat{\beta}, s_{\text{init}}, \rho_k, w_k$}
    \State $\mathbf{W} \leftarrow \text{diag}(w_k(r(\hat{\beta}) / s))$
    \State $\hat{\beta} \leftarrow (\mathbf{X}^\top \mathbf{W} \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{W} y$
    \State $r(\hat{\beta}) \leftarrow y - \mathbf{X} \hat{\beta}$
    \State $s \leftarrow \Call{S-scale-iteration}{r(\hat{\beta}), s_{\text{init}}, \, \rho_{k}(x)}$
    \State \Return $(\hat{\beta}, s)$
\EndFunction

\end{algorithmic}
\end{algorithm}

## Normal distribution of MM-estimator slopes {#sec-m-est-normal-dist}

First we check a histogram and QQplot of the MM-estimator slopes in @fig-slope-histogram-appendix. Visually it appears the slopes follow an approximately normal distribution. Then in @tbl-mm-ad-test-appendix we use an Anderson-Darling (AD) test which fails to reject the null hypothesis of normality. Taking these together there is strong evidence the MM-estimator estimates follow a normal distribution.

```{r}
#| label: fig-slope-histogram-appendix
#| fig-cap: Histogram of slopes (above) and QQplot (below) of MM-estimator simulated in @sec-normal-eff-sim for the $N=500$ case. The QQplot is included as an additional check despite the histogram appearing to have the shape of a normal distribution.
#| fig-width: 6
#| fig-height: 3

df <- read.csv("../data/normal_efficiency.csv")

hist <- df %>%
    filter(N_pts == 500) %>%
    ggplot(aes(x = slope_mm)) +
    geom_histogram(bins = 30) +
    labs(x = TeX("$\\hat{\\beta}_{MM}$"), y = "Count") +
    theme_minimal()

qqplot <- df %>%
    filter(N_pts == 500) %>%
    ggplot(aes(sample = slope_mm)) +
    stat_qq(size = 0.15) +
    stat_qq_line() +
    theme_minimal()

hist / qqplot
```

```{r}
#| label: tbl-mm-ad-test-appendix
#| tbl-cap: Anderson-Darling (AD) test p-values by $N$.

df %>%
    group_by(N_pts) %>%
    summarise(ad_pval = ad.test(slope_mm)$p.value) %>%
    mutate(ad_pval = round(ad_pval, 3)) %>%
    kbl(
        col.names = c("$N$", "AD p-value"),
        format = "latex",
        booktabs = TRUE,
        escape = FALSE,
        align = "rc"
    ) %>%
    kable_styling(latex_options = c("hold_position"))
```

## Breakdown of the intercept {#sec-breakdown-intercept-appendix}

We show details of the intercept term in @tbl-breakdown-intercept-appendix, from the simulation in @sec-breakdown-sim. We don't see a decrease of any metric after the breakdown point, although this is due to the choice of contamination being $y=10000$. It seems that the breakdown of the intercept occurs earlier than that for the slope, results could be simulated over more values of $N$ and $\epsilon$ with different contamination types to verify this further. 

```{r}
#| label: tbl-breakdown-intercept-appendix
#| tbl-cap: Simulation of breakdown of intercept from MM-estimator from @sec-breakdown-sim. @tbl-breakdown-slope shows the results for the slope estimate.

df <- read.csv("../data/breakdown_200.csv")

df %>%
    group_by(cp) %>%
    summarise(
        OLS_rmse = rmse(intercept_ols, 0),
        MM_rmse = rmse(intercept_mm, 0),
        OLS_var = var(intercept_ols),
        MM_var = var(intercept_mm),
        OLS_bias = mean(intercept_ols) - 0,
        MM_bias = mean(intercept_mm) - 0
    ) %>%
    mutate() %>%
    kbl(
        col.names = c(
            "$\\epsilon$",
            "$\\text{RMSE}(\\hat{\\beta}_{\\text{OLS}})$",
            "$\\text{RMSE}(\\hat{\\beta}_{\\text{MM}})$",
            "$\\operatorname{Var}({\\hat{\\beta}_{\\text{OLS}}})$",
            "$\\operatorname{Var}({\\hat{\\beta}_{\\text{MM}}})$",
            "$\\text{Bias}({\\hat{\\beta}_{\\text{OLS}}})$",
            "$\\text{Bias}({\\hat{\\beta}_{\\text{MM}}})$"
        ),
        format = "latex",
        booktabs = TRUE,
        escape = FALSE,
    ) %>%
    kable_styling(latex_options = c("hold_position"))
```

## MAD of the normal distribution {#sec-normal-dist-mad-proof}

We provide a proof of the fact that

$$ \text{MAD}(X)=\sigma\Phi^{-1}(0.75) \quad\text{with}\quad X\sim\mathcal{N}(\mu,\sigma^2). $$

Noting that the median of the normal distribution is $\mu$, we find for $X\sim\mathcal{N}(\mu, \sigma^2)$ 

\begin{align*}
\text{MAD}(X) &= \text{median}|X-\text{median}(X)|\\
    &= \text{median}|X-\mu|
\end{align*}

Using the idea that the median divides the sample in half, we can obtain

\begin{align*}
\frac{1}{2} &= \mathbb{P}(|X-\mu| \leq \text{MAD})\\
    &= \mathbb{P}(|Z| \leq \text{MAD}/ \sigma)\\
    &= \mathbb{P}(-\text{MAD}/ \sigma \leq Z \leq \text{MAD}/ \sigma)\\
    &= \Phi(\text{MAD}/ \sigma) - \Phi(-\text{MAD}/ \sigma)\\
    &= 2\Phi(\text{MAD}/ \sigma) - 1\\
\frac{3}{4}    &= \Phi(\text{MAD}/ \sigma)\\
\Phi^{-1}\left(\frac{3}{4}\right)&=\text{MAD}/ \sigma\\
\implies \text{MAD} &= \sigma \Phi^{-1}\left(\frac{3}{4}\right) \approx 0.6745 \sigma
\end{align*}


## Theil-Sen estimator {#sec-theil-sen}

The Theil-Sen (TS) linear regression estimator is defined as the median of the slopes

$$
\underset{\substack{i, j \\ j>i}}{\text{median}}\left(\frac{y_j-y_i}{x_j-x_i}\right)
$$

and it can be shown this estimator has a breakdown point of $\approx 29.3$%, see @huber_robust_stats[p. 66]. It suffers from low normal efficiency like S-estimators, but does not generalise to higher dimensions. A naïve implementation runs in $\mathcal{O}(n^2)$ time, however the TS formula is very similar to calculation of Kendall's $\tau$, which allows for computation in $\mathcal{O}(n\log n)$ time using Merge Sort (see @tau_nlogn).